---
title: "P8105 HW2"
author: "Mahdi Maktabi - mm6410"
date: 2024-10-01
output: github_document
---

```{r setup, echo = FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
```

This document is P8105 Homework 2.

I loaded necessary packages (e.g. `tidyverse` and `dplyr`). 

## Problem 1

```{r}
subway_df = read_csv(
  file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  na = c(".", "NA", "")) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

The data has 1,868 observations and 19 variables. The variables include line, station name, station longitude/latitude, routes 1-11, entry and entrance type, vending, and ADA compliance. 

To clean the data, the CSV data file was imported, missing values were excluded, and the variable names were cleaned such that they were all lowercase. I selected only the variables listed above and changed the "entry" column so that YES/NO character values were converted to TRUE/FALSE logistic values.



